Grade Retention and Mixed Race

### 2022-09-02 products/paper/main.qmd
Write rough draft of prior work section

-   Review previous studies on outcomes

    -   Campbell
    -   Herman
    -   Holloway/Ellis stuff
    -   Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
    -   Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
    -   Limitations of this work
        -   Small sample sizes lead to inconsistent estimates in most of the survey data.
        -   In many cases, impossible to understand outcomes for part-Latino biracials.

>>>>

Prior work on the outcomes for biracial individuals has been hampered by several methodological issues. With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents. We know that far fewer people identify as biracial than could feasibly do so based on their parents' races (CITATION) and that various selection effects operate on the decision to identify as biracial among hose who could (CITE DAVENPORT, ROTH, OTHERS). Furthermore, data on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves (CITE SAPERSTEIN AND CO). For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination. Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups (CITE BRATTER).

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. CITE CAMPBELL used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. CITE HERMAN used data on high school students in California and Wisconsin in which the students reported their parents' races to examine differences in GPA. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. CITE KAO uses data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.

The primary difficulty with the survey data used in the preceding studies is that the sample size for most biracial groups will be small and thus results tend to be imprecise and inconclusive. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources includes relatively little detailed information. This approach has been used previously to examine the likelihood of living in poverty (CITATIONS) and residential segregation (CITATIONS). SUMMARIZE FINDINGS. While these findings provide valuable information of the context of biracial childrens' lived experience, but do not directly measure their own outcomes, but rather the experiences of their parents.

However, one important educational outcome is identifiable for children in Census and ACS data - grade retention. Grade retention is partially identifiable in these data sources and provides one of the earliest measurable forms of stratification that individuals face in their lives.
